---
title: "6_go_ann"
author: "zhengchen"
date: "2024/6/29"
output: html_document
---
# 1 prepare work
### 1.1 load packages and data
```{r message=FALSE,warning=FALSE}
# rm(list=ls())
library(stringr)
library(tidyr)
library(tibble)
library(dplyr)
library(readr)
library(plyr)
library(readxl)
library(ggplot2)
library(paletteer)
library(Rmisc)
library(cowplot)
library(ggpubr)
library(gplots)
library(goseq)
library(clusterProfiler)
library(topGO)
library(GO.db)
library(reshape2)
#install.packages("R.utils")
R.utils::setOption("clusterProfiler.download.method",'auto') 
#browseVignettes("goseq")
library('org.Hs.eg.db')
load("../medianrdata/snv.Rdata")
load("../medianrdata/snvlist.Rdata")
load("../medianrdata/num_snv_107.Rdata")
```

# 2. Data organization
The goseq analysis object contains two sets of information: 1. gene ID; 2. whether it is differentially expressed. Among all reference genes, the presence of somatic mutations is considered differentially expressed and assigned the value T, and the absence of somatic mutations is assigned the value F.
Consider mutation characteristics: 1. Selection of non-synonymous mutations
```{r message=FALSE,warning=FALSE}
outline=c("P1-4","P6-37","P13L-3","P15L-5","P21L-8","P21N-1","P21N-5")
paired=c("P13L","P13N","P15L","P15N","P21L","P21N","P7L","P7N")
paired_sc=num_snv$sc[num_snv$patient %in% paired]
#view(snv_vcf)
snv_vcf$group=num_snv$group[match(snv_vcf$sample,num_snv$sc)]
snv_vcf=filter(snv_vcf,!sample %in% outline)
gene_table=table(snv_vcf$Gene_ID,snv_vcf$group,snv_vcf$sample) %>% as.data.frame()
#gene_table=filter(gene_table,Freq!=0)
gene_table=gene_table[str_length(gene_table$Var1)==15,]

# In all ref gene, mutations equal to there is an enrichment
refgne=read.delim("../infodata/gene_RefSeqGene.txt")
#refgne=refgne[!duplicated(refgne$Gene.stable.ID),]
#转ENSEMBLE ID
ref_ID= bitr(refgne$GeneID,fromType = 'ENTREZID',toType = 'ENSEMBL',OrgDb = 'org.Hs.eg.db')
ref_ID=ref_ID[!duplicated(ref_ID$ENTREZID),]
# refgne$GeneID=as.character(refgne$GeneID)
# refgne2=right_join(EN_ID,refgne,by=c("ENTREZID"="GeneID"))


table(gene_table$Var2)
gene_OAlesion=ref_ID$ENSEMBL %in% gene_table$Var1[gene_table$Var2=="OA lesion" & gene_table$Freq!=0]
gene_OAnonlesion=ref_ID$ENSEMBL %in% gene_table$Var1[gene_table$Var2=="OA non-lesion"& gene_table$Freq!=0]

gene_OAlesionpair=ref_ID$ENSEMBL %in% gene_table$Var1[gene_table$Var2=="OA lesion" & gene_table$Freq!=0&gene_table$Var3 %in% paired_sc]
gene_nonOAcontrol=ref_ID$ENSEMBL %in% gene_table$Var1[gene_table$Var2=="non-OA control"& gene_table$Freq!=0]

#table(gene_OAlesion)
names(gene_OAlesion)=ref_ID$ENTREZID
names(gene_OAlesionpair)=ref_ID$ENTREZID
names(gene_OAnonlesion)=ref_ID$ENTREZID
names(gene_nonOAcontrol)=ref_ID$ENTREZID

```

# 3.GO analysis
```{r fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
#go分析
my_goseq=function(gene_x,raw=FALSE){
  pwf=nullp(gene_x,"hg19","knownGene",plot.fit = F)
  GO=goseq(pwf,"hg19","knownGene",method = "Wallenius")
  GO$over_represented_pvalue_corrected=p.adjust(GO$over_represented_pvalue,method="fdr")
  GO_filtered=GO[#GO$over_represented_pvalue<0.05&
                   GO$numDEInCat>=10&
                   GO$numInCat<=1000,]
  if (!raw) {
  GO_filtered  
  }
  else {
  GO
  }
}

go_OAlesion=my_goseq(gene_OAlesion)
go_OAlesion$group="OA lesion"

go_OAlesionpair=my_goseq(gene_OAlesionpair)
go_OAlesionpair$group="OA lesion paired"

go_nonOAcontrol=my_goseq(gene_nonOAcontrol)
go_nonOAcontrol$group="non-OA control"

go_OAnonlesion=my_goseq(gene_OAnonlesion)
go_OAnonlesion$group="OA non-lesion"
```
#3.1 Test for goterm with differences between groups
```{r message=FALSE,warning=FALSE}
my.indexDiff <- function(go,input1,input2)
{
	a=which(input1$category==go)/nrow(input1)
	b=which(input2$category==go)/nrow(input2)
	if(length(a)==0)
	{
		a=1
	}
	if(length(b)==0)
	{
		b=1
	}
	a-b
}


#permutation test
permutations_test=function(gene_x,gene_y,go_x,go_y,iteration){

  go_x_list=as.character(go_x$category[go_x$over_represented_pvalue<0.01])
  go_y_list=as.character(go_y$category[go_y$over_represented_pvalue<0.01])
  go_list=unique(c(go_x_list,go_y_list))
  idx_diff_permutation=numeric(0)

  if (iteration>1) {
  for (i in 1:iteration) {
  gene_x=sample(gene_x)
  gene_y=sample(gene_y)
  names(gene_x)=ref_ID$ENTREZID
  names(gene_y)=ref_ID$ENTREZID
  go_x=my_goseq(gene_x)
  go_y=my_goseq(gene_y)
  idx_diff_permutation=rbind(idx_diff_permutation,
                             sapply(go_list,function(x){my.indexDiff(x,go_x,go_y)})) 
  print(paste0(i,"/",iteration))}  
  }
  else {
  names(gene_x)=ref_ID$ENTREZID
  names(gene_y)=ref_ID$ENTREZID
  idx_diff_permutation=rbind(idx_diff_permutation,
                             sapply(go_list,function(x){my.indexDiff(x,go_x,go_y)}))     
  }
  
  as.data.frame(idx_diff_permutation)
  
}

idx_diff=permutations_test(gene_OAlesion,gene_nonOAcontrol,go_OAlesion,go_nonOAcontrol,1)
idx_diff_permutation=permutations_test(gene_OAlesion,gene_nonOAcontrol,go_OAlesion,go_nonOAcontrol,1000)

idx_diff_pair=permutations_test(gene_OAlesionpair,gene_OAnonlesion,go_OAlesion,go_OAnonlesion,1)
idx_diff_permutation_pair=permutations_test(gene_OAlesionpair,gene_OAnonlesion,go_OAlesion,go_OAnonlesion,1000)

save(idx_diff,idx_diff_permutation,file = "../medianrdata/pernull.Rdata")
save(idx_diff_pair,idx_diff_permutation_pair,file = "../medianrdata/pernull_pair.Rdata")


gen_gotable=function(gene_x,gene_y){
  go_x=my_goseq(gene_x)
  go_x_raw=my_goseq(gene_x,raw = T)
  go_y=my_goseq(gene_y)
  go_y_raw=my_goseq(gene_y,raw = T)
  
  gene_x_list=names(gene_x)[gene_x]
  gene_y_list=names(gene_y)[gene_y]  
  
  go_x_list=as.character(go_x$category[go_x$over_represented_pvalue<0.01])
  go_y_list=as.character(go_y$category[go_y$over_represented_pvalue<0.01])
  
  go_list=unique(c(go_x_list,go_y_list))
  gene_list=unique(c(names(gene_x)[gene_x],names(gene_y)[gene_y]))
  gene_map=getgo(gene_list,"hg19","knownGene")
  
  go_table=data.frame(go_list,

					 sapply(go_list,function(x){sum(gene_x_list %in% names(gene_map)[grep(x,gene_map)],na.rm=T)}),
					 sapply(go_list,function(x){sum(gene_y_list %in% names(gene_map)[grep(x,gene_map)],na.rm=T)}),
					 sum(gene_x,na.rm=T),
					 sum(gene_y,na.rm=T),
					 sapply(go_list,function(x){go_x_raw$numDEInCat[go_x_raw$category==x]}),
					 sapply(go_list,function(x){go_y_raw$numDEInCat[go_y_raw$category==x]}),
					 sapply(go_list,function(x){go_x_raw$numInCat[go_x_raw$category==x]}),
					 sapply(go_list,function(x){go_x_raw$over_represented_pvalue[go_x_raw$category==x]}),
					 sapply(go_list,function(x){go_y_raw$over_represented_pvalue[go_y_raw$category==x]}),
					 sapply(go_list,function(x){go_x_raw$term[go_x_raw$category==x]}),
					 sapply(go_list,function(x){go_x_raw$ontology[go_x_raw$category==x]})
					)
  
   colnames(go_table)=c("category","hit_x","hit_y","total_x","total_y","numDE_x","numDE_y","numInCat","p_x","p_y","term","ontology")
   go_table$corrected_p_x=p.adjust(go_table$p_x,method="fdr")
   go_table$corrected_p_y=p.adjust(go_table$p_y,method="fdr")
  go_table
}

go_table=gen_gotable(gene_OAlesion,gene_nonOAcontrol)
go_table_pair=gen_gotable(gene_OAlesionpair,gene_OAnonlesion)

save(go_table,file = "../medianrdata/gotable.Rdata")
save(go_table_pair,file = "../medianrdata/gotable_pair.Rdata")
```
# 3.2 Selection of differential genes
```{r}
load(file = "../medianrdata/gotable.Rdata")
load(file = "../medianrdata/gotable_pair.Rdata")
load(file = "../medianrdata/pernull.Rdata")
load(file = "../medianrdata/pernull_pair.Rdata")

go_permutation=function(go_table,idx_diff,idx_diff_permutation){
  go_table$p_permutation=sapply(1:nrow(go_table),
                              function(x){
                                if(idx_diff[x]<0){sum(idx_diff[,x]>idx_diff_permutation[,x])/length(idx_diff_permutation[,x])}
                                else{sum(idx_diff[,x]<idx_diff_permutation[,x])/length(idx_diff_permutation[,x])}})
go_table$corrected_p_permutation=p.adjust(go_table$p_permutation,method="fdr")
#go_table$annotation=GOTERM[[go_table$category]]
#write.table(go_table,file="GO.tsv",quote=F,sep="\t",row.names=F)
go_table
}

go_table=go_permutation(go_table,idx_diff,idx_diff_permutation)
go_table_pair=go_permutation(go_table_pair,idx_diff_pair,idx_diff_permutation_pair)


plot_go_table=function(go_table,name_x,name_y,pos="none",warp=30){
  go_table$p_deviance=-log10(go_table$corrected_p_x)+log10(go_table$corrected_p_y)
  ready=go_table[go_table$corrected_p_x<0.05&go_table$corrected_p_y<0.05&go_table$p_permutation<0.01,c(1,11,12,13,14)] %>% melt()
  colnames(ready)=c("GO","Term","ontology","Variable","Pvalue")
  ready$p_devi=go_table$p_deviance[match(ready$Term,go_table$term)]
  ready$Term=factor(ready$Term,levels=unique(ready$Term[order(ready$p_devi,decreasing = F)]))
  ready$Tissue=factor(sapply(ready$Variable,function(x){if(grepl("x",x)){name_x}else if(grepl("y",x)){name_y}}),levels=c(name_y,name_x))
  ready$Pconvert=-log10(ready$Pvalue)
  color = c("#F8766D","#619CFF","#00BA38")
  names(color)=c("OA lesion","OA non-lesion","non-OA control")
ggplot() +
     geom_col(data=ready,aes(x=Term,y=Pconvert,fill=Tissue),position="dodge") +
     geom_hline(yintercept=-log10(0.05),linetype="dashed") +
     scale_fill_manual(values = color[names(color) %in% c(name_y,name_x) ])+
     xlab("") + ylab("-log10(FDR-adjusted P-value)") +
     scale_y_continuous(expand = c(0,0),limits = c(0,8),breaks = seq(0,100,2))+
     scale_x_discrete(labels=function(x) str_wrap(x, width=warp))+
     #ggtitle(label = plot_name,)+
     coord_flip() +theme_classic()+ 
     theme(legend.position=pos,
           legend.title = element_blank(),
           legend.box.margin = margin(t=0,r=0,b=-25,l=-40,"pt"),
           legend.text = element_text(size = 6,family = "sans"),
           legend.key.height = unit(6,"pt"),
           legend.key.width = unit(6,"pt"),
           axis.text.x = element_text(size = 7,family = "sans"),
           axis.text.y = element_text(size = 7,family = "sans"),
           axis.title.x = element_text(size = 7,family = "sans",hjust = -0.2))
}

#color = c("#F8766D","#00BA38","#619CFF")
#names(color)=c("nonOA control","OA lesion","OA nonlesion")

p1=plot_go_table(go_table,"OA lesion","non-OA control",pos =c(0.95,0.99),warp=50)
p2=plot_go_table(go_table_pair,"OA lesion","OA non-lesion",pos =c(0.95,0.99),warp=20)

t1=go_table[go_table$corrected_p_x<0.05&go_table$corrected_p_y<0.05&go_table$p_permutation<0.01,c(1,11,12,13,14,15,16)]
t2=go_table_pair[go_table_pair$corrected_p_x<0.05&go_table_pair$corrected_p_y<0.05&go_table_pair$p_permutation<0.01,c(1,11,12,13,14,15,16)]

# ggsave("../result/49_go_OAlesion.tiff", plot = p1 ,units = "in", width =5, height =3.5, dpi = 600, compression = 'lzw')
# ggsave("../result/50_go_OAlesion.tiff", plot = p2 ,units = "in", width =5, height =10, dpi = 600, compression = 'lzw')

ggsave("../manuscript7/P6/go_1.pdf", plot = p1 ,units = "in", width =4.5, height =2.5)
ggsave("../manuscript7/P6/go_2.pdf", plot = p2 ,units = "in", width =2.7, height =7.5)

write.csv(t1,file = "../manuscript7/go_table1.csv",row.names = F)
write.csv(t2,file = "../manuscript7/go_table_paired2.csv",row.names = F)
```


# 4.plot
```{r}
plot_top10=function(go_x,plot_name,pos){
   ready=go_x[,c(1,2,6:8)]
   colnames(ready)=c("GO","Pvalue","Term","ontology","group")
   ready$Pconvert=-log10(ready$Pvalue)
   a=ready[order(ready$Pconvert,decreasing =T),] %>% filter(ontology=="BP") %>% head(5)
   b=ready[order(ready$Pconvert,decreasing =T),] %>% filter(ontology=="CC") %>% head(5)
   c=ready[order(ready$Pconvert,decreasing =T),] %>% filter(ontology=="MF") %>% head(5)

   a$Term=factor(a$Term,levels = a$Term[order(a$Pconvert)])
   b$Term=factor(b$Term,levels = b$Term[order(b$Pconvert)])
   c$Term=factor(c$Term,levels = c$Term[order(c$Pconvert)])
   
   d=rbind(c,a,b)
   d$ontology=factor(d$ontology,levels = c("CC","BP","MF"))
   
   p=ggplot() +
     geom_col(data=d,aes(x=Term,y=Pconvert,fill= ontology),width = 0.75) +
     geom_hline(yintercept=-log10(0.05),linetype="dashed") +
     scale_fill_discrete(name="") + xlab("") + ylab("-log10(FDR-adjusted P-value)") +
     scale_y_continuous(expand = c(0,0),limits = c(0,11),breaks = seq(0,100,5))+
     scale_x_discrete(labels=function(x) str_wrap(x, width=30))+
     scale_fill_manual(values = c("#DDAA33FF","#008348FF","#E13A3EFF"))+
     ggtitle(label = plot_name,)+
     coord_flip() +theme_classic()+ 
     theme(legend.position=pos,
           plot.title = element_text(hjust = 0.5),
           axis.text.x = element_text(size = 11),
           axis.text.y = element_text(size = 11),
           )
   
   for (go in d$GO) {
     print(GOTERM[[go]])
     # print(GOTERM[[go]]@Ontology)
     # print(GOTERM[[go]]@Definition)
      cat("--------------------------------------\n")
   }
   p
}


```

# 5.go_inter_plot

```{r}
#intersection
go_inter_plot=function(go_x,go_y,go_z){
  go_x$Pconvert=-log10(go_x$over_represented_pvalue)
  go_y$Pconvert=-log10(go_y$over_represented_pvalue)
  go_z$Pconvert=-log10(go_z$over_represented_pvalue)
  
  
  x=go_x[,c("category","Pconvert")]
  y=go_y[,c("category","Pconvert")]
  z=go_z[,c("category","ontology","Pconvert")]
  a=inner_join(x,y,by= "category")
  s=inner_join(z,a,by= "category")
  s$ps=rowSums(s[,3:5])
  
 
plot_onto=function(X){
  inner=s$category[s$ontology==X] %>%  head(10)
  inall=rbind(
    filter(go_x, category %in% inner),
    filter(go_y, category %in% inner),
    filter(go_z, category %in% inner)
  )[,-c(2:5)]

     for (go in inall$category) {
     print(GOTERM[[go]])
     # print(GOTERM[[go]]@Ontology)
     # print(GOTERM[[go]]@Definition)
      cat("--------------------------------------\n")
   }   
  ggplot() +
     geom_col(data=inall,aes(x=term,y=Pconvert,fill= group),position="dodge") +
     geom_hline(yintercept=-log10(0.05),linetype="dashed") +
     scale_fill_discrete(name="") + xlab("") + ylab("-log10(FDR-adjusted P-value)") +
     scale_y_continuous(expand = c(0,0),limits = c(0,25),breaks = seq(0,25,1))+
     scale_x_discrete(labels=function(x) str_wrap(x, width=30))+
     #ggtitle(label = plot_name,)+
     #facet_wrap(~ontology)+
     coord_flip() +theme_classic()+
     theme(legend.position="bottom",plot.title = element_text(hjust = 0.5))  
  

  }

print(s)

CC=plot_onto("CC")   
BP=plot_onto("BP")
MF=plot_onto("MF")

plot_grid(CC,BP,MF,
          ncol = 3,labels = c("CC","BP","MF"))

  
}



```

# 6.plot

```{r fig.height=4, fig.width=13}
g1=plot_top10(go_OAlesion,"OA lesion",pos = "none")
g2=plot_top10(go_OAnonlesion,"OA non-lesion","none")
g3=plot_top10(go_nonOAcontrol,"non-OA control",pos = c(0.8,0.15))

goterm=sapply(unique(go_table$category,go_table_pair$category),function(x){GOTERM[[x]]})
goterm_t=sapply(names(goterm),function(x){rbind(goterm[[x]]@GOID,
                                       goterm[[x]]@Term,
                                       goterm[[x]]@Ontology,
                                       goterm[[x]]@Definition)}) %>% t()

colnames(goterm_t)=c("GOID","Term","Ontology","Definition")

g1=filter(go_OAlesion,over_represented_pvalue_corrected<0.05)
g2=filter(go_OAnonlesion,over_represented_pvalue_corrected<0.05)
g3=filter(go_nonOAcontrol,over_represented_pvalue_corrected<0.05)

```


